---
title: "Factors"
author: "Ethan Pisterman"
date: "2024-10-24"
output: html_document
---
# Introduction

This document demonstrates data transformations and visualizations applied to the Rice Center dataset. We reformat weekdays, classify day types (Weekday vs. Weekend), assign lunar phases, and assess tidal movements.

## Loading Libraries and Data
```{r setup, message=FALSE, warning=FALSE}
library(lubridate)
library(dplyr)
library(forcats)
library(ggplot2)

source("/Users/ethanpisterman/Desktop/ENVS  543/ENVS543/get_rice_data.R")
r_data <- get_rice_data()
```

## Reorder weekdays to start on Monday
```{r}
r_data$Weekday <- fct_relevel(r_data$Weekday,
                              c("Monday", "Tuesday", "Wednesday",
                                "Thursday", "Friday", "Saturday", "Sunday"))
# Removing NA values
r_data_filtered <- r_data %>% filter(!is.na(Weekday))
# Bar Chart
ggplot(r_data_filtered, aes(x = Weekday)) +
  geom_bar() +
  theme_minimal()
```

## Adding Weekday/Weekend Classification
```{r}
# Weekday or Weekend column
r_data_filtered <- r_data_filtered %>%
  mutate(day_end = ifelse(Weekday %in% c("Saturday", "Sunday"), 
                                  "Weekend", "Weekday"))
# Convert to ordered factor
r_data_filtered$day_end <- factor(r_data_filtered$day_end, 
                                 levels = c("Weekday", "Weekend"),
                         ordered = TRUE)
ggplot(r_data_filtered, aes(x = Weekday,
                            fill = day_end)) +
  geom_bar(position = "dodge") +
  theme_minimal() +
  labs(title = "Distribution of Observations by Weekday and Weekend",
       x = "Day of the Week",
       y = "Count")
```

## Assigning Lunar Phases
```{r}
# Moon phase column
lunar_phase <- function(date) {
  day_in_cycle <- as.numeric(difftime(date,
                                      as.Date("2024-01-01"),
                                      units = "days")) %% 29.53
  case_when(
    day_in_cycle < 7.4 ~ "New Moon",
    day_in_cycle < 14.8 ~ "First Quarter",
    day_in_cycle < 22.1 ~ "Full Moon",
    TRUE ~ "Last Quarter")
}

r_data_filtered <- r_data_filtered %>%
  mutate(lunar_phase = lunar_phase(Date))

ggplot(r_data_filtered, aes(x = lunar_phase)) +
  geom_bar() +
  theme_minimal() +
  labs(title = "Distribution by Lunar Phase", x = "Lunar Phase", y = "Count")
```

## Determining Tide Direction
```{r}
# Tidal
r_data_filtered <- r_data_filtered %>%
  arrange(DateTime) %>%
  mutate(tide_direction = ifelse(
    lead(SurfaceWaterElev_m_levelNad83m) > SurfaceWaterElev_m_levelNad83m, 
    "Incoming", 
    "Outgoing"
  ))
ggplot(r_data_filtered, aes(x = DateTime, y = SurfaceWaterElev_m_levelNad83m, color = tide_direction)) +
  geom_line() +
  labs(title = "Tide Direction Over Time",
       x = "DateTime",
       y = "Surface Water Elevation (m)",
       color = "Tide Direction") +
  theme_minimal()
```

